Ordinal regression models made easy. A tutorial on parameter interpretation, data simulation, and power analysis.

### 2024-01-29 paper/paper.tex
reducing words

\abstract{%
Ordinal data are very common in psychology such as Likert items, ratings or generic ordinal variables. Despite the widespread these variables are usually analyzed using metric models (e.g., standard linear regression). This approach can have important drawbacks in terms of statistical inference (e.g., reduced power) and prediction. One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis. The aim of the tutorial is presenting ordinal regression models using a simulation-based approach. In the first section we introduced the general model highlighting the crucial components and assumptions. Then we explained how to intepret the parameters for a logit and probit model. Then we proposed two way for simulating ordinal data as a function of predictors showing common research design such as a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor. Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors. The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models. The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Github repository.
}

@@

In contrast to nominal scales, the labels in ordinal scales are ordered. Unlike interval or ratio scales, there is no explicit assumption about the distance between labels. An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement). Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer. Stevens (1946) and Kemp and Grace (2021) suggested that for ordinal variables is appropriate to calculate ranks-based descriptive statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation) and using appropriate inferential tools (Agresti, 2010; e.g., N. Cliff, 1996). This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data. Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels (Liddell \& Kruschke, 2018). More generally, Norman Cliff (2016) suggests that several research questions in behavioral sciences can be considered as ordinal (\emph{is the score} \(x\) higher than the score \(y\)?) concerning variables where the most appropriate measurement scale is probably ordinal.

In Psychology especially when using item-based measures (questionnaires, surveys, etc.) the common practice is using a linear regression that makes an explicit assumption about metric features of the response variable. Liddell and Kruschke (2018) reviewed the psychological literature using Likert-based measures and reported how the majority of papers used metric-based statistical models. In the same work, Liddell and Kruschke (2018) showed examples about the potential pitfalls of metric models applied to ordinal variables (but see Robitzsch, 2020 for an alternative perspective). They reported problems in terms of lack of power, inversion of the effects (e.g., finding a negative effect when the true effect is positive) and biased estimates. While sums or averages of ordinal items could be treated as metric (Carifio \& Perla, 2008; Carifio \& Perla, 2007; but see Jamieson, 2004) Liddell and Kruschke (2018) provided some examples of potential pitfalls even in this case. In the current paper we discuss only cases where there is a single ordinal outcome (e.g., a single item, question, etc.) without considering the case of aggregating ordinal responses.

@@

@@

Despite the specific model proposed by Liddell and Kruschke (2018) proposed a specific modeling approach, there is a class of regression models that consider the ordinal nature of the response variable. The actual statistical nomenclature of \emph{ordinal regression} models can be confusing (Tutz, 2022). Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.

We can identify three main types: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990), and \emph{adjacent category models}. Among these, the cumulative model is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds. The \emph{sequential model} is appropriate when modeling sequential processes. Assuming to have five response options, the sequential model assume that responding ``3'' assume a sequential process where steps ``1'' and ``2'' are already reached. A clear example is proposed by Bürkner and Vuorre (2019) where the marriage duration in years is predicted as a function of some explanatory variables. For each level of the response variable there is a latent distribution where the step between a marriage year \(k = 1\) and the next years \(k > 1\) is modeled by the sequential model. When comparing \(k\) with \(k > 1\), everything lower than \(k\) is assumed to be already reached (Tutz \& Berger, 2020).

The adjacent category model compare the category \(k\) with \(k + 1\) still assuming a latent distribution for each \(k\). As suggested by Tutz (2022) the adjacent-category model can be seen as a series of binary binomial regressions taking into account the order of the categories. Bürkner and Vuorre (2019) suggested that adjacent-category model can be chosen for its mathematical convenience and there is no a clear empirical interpretation as for the cumulative vs sequential model.

In the current paper we put the focus on the CM for several reasons. Firstly, the latent formulation of the model is particularly convenient for parameter interpretation and data simulation. The second reason is that several psychological variables can be formalized as a latent continuous variable observed as an ordinal item. Furthermore, the signal detection theory framework that is very common in experimental psychology can be implemented using a CM (e.g., DeCarlo, 2010). Figure \ref{fig:fig-explain-cumulative} depicts the overall structure of the CM.

@@

In this section we introduce some notation for the CM to be consistent with the existing literature and the \texttt{ordinal} R package (Christensen, 2022). We define \(Y_k\) as the observed ordinal variable with \(k\) levels and \(Y^\star\) is the underlying latent variable. The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \dots, \alpha_{k - 1}\). Similarly to the generalized linear models framework (e.g., Fox, 2015), we define \(g(x) = \eta\) as the link function that maps probabilities into the linear predictor \(\eta\). To transform back \(\eta\) into probabilities we use the inverse of the link function \(x = g^{-1}(\eta)\). The specific link function define the type of model and require a different R function. For example, when assuming a Gaussian distribution (\emph{probit} link function) entails using the cumulative distribution function \(g(x) = \Phi^{-1}(x) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(x = g^{-1}(\eta) = \Phi(\eta)\).

When modelling an ordinal variable in a cumulative link model we actually modelling the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(\mathbf{X}\) value \(P(Y \leq k | \mathbf{X}_i)\)}. Equation \eqref{eq:prob-cum-model1} shows the general cumulative model including predictors \(\mathbf{X}\) and regression coefficients \(\boldsymbol{\beta}\). The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta_j\) as in the standard regression models where higher \(\beta\) values corresponds to increased probability of responding higher \(k\) categories (Agresti, 2010).

@@

The \(\mathbf{X} \boldsymbol{\beta}\) is the linear predictor \(\eta\) that is the cumulative probability \(P(Y \leq k)\) transformed using the link function \(g(\cdot)\). To obtain the probability of a single outcome \(P(Y = k)\) we can compute the difference between cumulative probabilities as shown in Equation \eqref{eq:prob-cum-model2}.

@@

There are always two special cases when computing the probability of a single outcome \(Y\) that is when \(Y = 1\) and \(Y = k\). In the first case the cumulative probability is calculated from \(-\infty\) that is the same as temporary assuming an \(\alpha_0 = -\infty\). In the second case (\(Y = 1\)) the probability is calculated as \(P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)\) that is the same as assuming a temporary threshold \(\alpha_k = +\infty\). The Figure \ref{fig:fig-explain-cumulative} shows how the single probabilities of the ordinal outcome are calculated from cumulative probabilities.

@@

\begin{equation}
\mathbf{Y^\star} = \mathbf{X}\boldsymbol{\beta} + \mathbf{\epsilon}
\label{eq:latent-model}
\end{equation}

The crucial part is \(\epsilon\) that is the random part (errors) of the model. For a \emph{probit} model, errors are sampled from a standard Normal distribution while for a \emph{logit} model from a standard logistic distribution. Following the notation by Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).

The thresholds \(\alpha_k\) are considered fixed and part of the measurement procedure (Liddell \& Kruschke, 2018; see the \emph{location-shift} models by Tutz, 2022, where thresholds vary as a function of predictors). The model can be also formalized in alternative ways. Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian version of the model or Gelman, Hill, and Vehtari (2020) for other thresholds parametrizations.

The scale (spread) of the latent variable is fixed to one for both \emph{probit} and \emph{logit} models. In the supplementary materials we showed how to implement a location-scale models including predictors on the scale parameter (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022).

@@

The CM implemented in Equations \eqref{eq:prob-cum-model1} and \eqref{eq:latent-model} requires specifying the link function \(g(\cdot)\) or the errors distribution \(\epsilon_i \sim D(\mu, \sigma^2)\) (\(D\) being a certain probability distribution). Among several available functions the \emph{logit} and \emph{probit} models are the most common. The \emph{logit} model use a \emph{logit} link function and a logistic distribution as latent variable. On the other side, the \emph{probit} model assume a latent Gaussian distribution using the inverse of the cumulative distribution (\(\Phi^{-1}\)) function as link function.

@@

In terms of parameters, both distributions can be defined with in terms of location \(\mu\) and scale \(s\). The standard normal distribution has \(\mu = 0\) and \(s = 1\). Furthermore the variance corresponds to the scale \(s^2 = \sigma^2 = 1\). The variance of the logistic distribution is \(\sigma^2 = \frac{s^2\pi^2}{3}\). The standard logistic distribution has \(\mu = 0\) and \(s^2 = 1\) thus the standard deviation simplified to \(\frac{\pi}{\sqrt{3}} \approx 1.81\). In practical terms, fixing \(\mu\) and \(s\) lead to an higher standard deviation for the logistic distribution.

@@

For fitting the cumulative models we used the \texttt{ordinal} package (Christensen, 2019). Despite the presence of other possibilities the \texttt{ordinal} package provide the most complete and intuitive way to implement the ordinal models. The syntax is very similar to standard linear models in R and default functions to calculate predictions, perform model comparison, extract relevant model information are implemented similarly to standard regression modelling.\footnote{For a very complete overview of the ordinal package see Christensen (2019) and the Package documentation \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf}}

@@

When fitting the model the crucial arguments are the \texttt{formula}, the \texttt{link} function and the \texttt{data}. More advanced arguments are the \texttt{nominal}, \texttt{scale} and \texttt{threshold}.

\begin{itemize}
\tightlist
\item
  \texttt{formula}: the formula \texttt{y\ \textasciitilde{}\ x} with the dependent variable and predictors.
\item
  \texttt{link}: is the link function. In this tutorial we consider only the \emph{logit} and \emph{probit} link but other link functions are available.
\item
  \texttt{data}: is the dataset with variables included in the \texttt{formula}
\item
  \texttt{nominal}: formula with predictors where the proportional odds assumption (See Section \ref{proportional-odds-assumption}) is relaxed (i.e., partial or non proportional odds)
\item
  \texttt{scale}: formula with predictors for the scale (standard deviation) parameter. This argument allow to fit a scale-location model (see supplementary materials). The main \texttt{formula} argument refers to predictors on the location parameter (i.e., the mean \(\mu\)).
\item
  \texttt{threshold}: different structures for estimating the thresholds. The default is \texttt{threshold\ =\ "flexible"} where \(k - 1\) threshold (where \(k\) is the number of ordinal levels for \(Y\)) are estimated.
\end{itemize}

We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections. Table \ref{tab:tab-dataset-example-latex} contains simulated data from \(n = 100\) participants rating the agreement about a certain item with \(k = 4\) ordered options. The participants are divided into two groups (\(x_a\) and \(x_b\)). We can fit a cumulative link model with \texttt{clm()} function and check the model summary.

@@

\begin{table}
\centering
\caption{\label{tab:tab-dataset-example-latex}Summary of the simulated dataset. For each group (a and b) we reported mean, median and standard deviation of the ordinal outcome (metric descriptive statistics). For each ordinal outcome (\(Y\)) we reported the frequency and (within parentheses) the probability.}
\centering
\begin{tabular}[t]{cccccccc}
\toprule
\textbf{Group} & \textbf{Mean} & \textbf{Median} & \textbf{SD} & \textbf{Y1} & \textbf{Y2} & \textbf{Y3} & \textbf{Y4}\\
\midrule
a & 2.62 & 3 & 1.1 & \textbf{10} (p = \textit{0.2}) & \textbf{13} (p = \textit{0.26}) & \textbf{13} (p = \textit{0.26}) & \textbf{14} (p = \textit{0.28})\\
b & 3.06 & 3 & 1.04 & \textbf{6} (p = \textit{0.12}) & \textbf{7} (p = \textit{0.14}) & \textbf{15} (p = \textit{0.3}) & \textbf{22} (p = \textit{0.44})\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
\FunctionTok{summary}\NormalTok{(fit)}
\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x}
\CommentTok{\#\textgreater{} data:    dat}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{}  link  threshold nobs logLik  AIC    niter max.grad cond.H }
\CommentTok{\#\textgreater{}  logit flexible  100  {-}131.80 271.60 4(0)  2.13e{-}11 1.7e+01}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Coefficients:}
\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)  }
\CommentTok{\#\textgreater{} xb   0.7556     0.3685    2.05   0.0403 *}
\CommentTok{\#\textgreater{} {-}{-}{-}}
\CommentTok{\#\textgreater{} Signif. codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Threshold coefficients:}
\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
\CommentTok{\#\textgreater{} 1|2  {-}1.3328     0.3126  {-}4.264}
\CommentTok{\#\textgreater{} 2|3  {-}0.2185     0.2703  {-}0.808}
\CommentTok{\#\textgreater{} 3|4   0.9767     0.2893   3.377}
\end{Highlighting}
\end{Shaded}

@@

Odds and odds ratios are clearly defined with \(k = 2\) outcomes. With an ordinal variable and a cumulative model we can use the cumulative odds ratio. With e.g.~\(k = 4\) outcomes we have \(k - 1\) odds ratios determined by the cumulative probability in terms of \(P(Y \leq 1), \dots, P(Y \leq k - 1)\) comparing the two groups.

For example the probability of responding \(Y \leq 1\) in the group ``a'' is 0.20 corresponding to an odds of 0.25. When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y \leq 1\) we obtain that the group ``a'' has 1.83 times the odds of responding \(Y \leq 1\) compared to the group ``b''.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# function to calculate odds}
\NormalTok{odds }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(p) p }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ p)}

\CommentTok{\# the dummy\_ord() is a custom function creating k {-} 1 dummy variable for the cumulative probabilities}

@@

\CommentTok{\# calculating the cumulative probability for k {-} 1 variables. }
\CommentTok{\# Taking the average of a series of 0{-}1 is the same as computing the proportion of 1s.}
\NormalTok{(group\_a }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
\CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
\CommentTok{\#\textgreater{}    0.20    0.46    0.72}
\NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
\CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
\CommentTok{\#\textgreater{}    0.12    0.26    0.56}

>>>>

\abstract{%
Ordinal data are very common in psychology such as Likert items, ratings or generic ordered variables. Despite the widespread, these variables are usually analyzed using metric models (e.g., standard linear regression). This approach can have important drawbacks in terms of statistical inference (e.g., reduced power and increased type-1 error) and prediction. One of the possible reason of not using ordinal regression models could be the difficulty in understanding parameters (e.g., odds ratios) or conduting a power analysis. The aim of the tutorial is presenting ordinal regression models using a simulation-based approach. In the first section we introduced the general model highlighting the crucial components and assumptions. Then we explained how to intepret the parameters for a logit and probit model. Then we proposed two way for simulating ordinal data as a function of predictors showing common research design such as a 2x2 interaction with categorical predictors and the interaction between a numeric and categorical predictor. Finally, we showed an example of power analysis using simulations that can be easily extended to complex models with multiple predictors. The tutorial is supported by a collection of custom R functions developed to simulate and understand ordinal regression models. The code to reproduce the proposed simulation, the custom R functions and other examples of ordinal regression models can be found on the online Github repository.
}

@@

In contrast to nominal scales, the labels in ordinal scales are ordered. Unlike interval or ratio scales, there is no explicit assumption about the distance between labels. An example is asking people the degree of agreement about a certain statement using a scale from 1 (no agreement) to 7 (total agreement). Answering 4 compared to 2 suggest an higher agreement but we cannot affirm that there is two times the agreement compared to the second answer. Stevens (1946) and Kemp and Grace (2021) suggested that for ordinal variables is appropriate to calculate ranks-based descriptive statistics (e.g., median or percentiles) instead of metric statistics (e.g., mean or standard deviation) and using appropriate inferential tools (Agresti, 2010; e.g., Cliff, 1996). This distinction in terms of the appropriateness of certain descriptive statistics is also relevant when modeling data. Treating ordinal data as metric refers to assuming the labels as actual integer numbers thus assuming a fixed and know distance between levels (Liddell \& Kruschke, 2018).

In Psychology especially with item-based measures (questionnaires, surveys, etc.) the common practice is running a linear regression that makes an explicit assumption about metric features of the response variable. Liddell and Kruschke (2018) reviewed the psychological literature using Likert-based measures and reported how the majority of papers used metric-based statistical models. In the same work, Liddell and Kruschke (2018) showed examples about the potential pitfalls of metric models applied to ordinal variables (but see Robitzsch, 2020 for an alternative perspective). They reported problems in terms of lack of power, inversion of the effects (e.g., finding a negative effect when the true effect is positive) and biased estimates. While sums or averages of ordinal items could be treated as metric (Carifio \& Perla, 2008; Carifio \& Perla, 2007; but see Jamieson, 2004) Liddell and Kruschke (2018) provided some examples of potential pitfalls even in this case. In the current paper we discuss only cases where there is a single ordinal outcome (e.g., a single item, question, etc.) without considering the case of aggregating ordinal responses.

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(dplyr)}
\FunctionTok{library}\NormalTok{(tidyr)}
\FunctionTok{library}\NormalTok{(ordinal)}
\FunctionTok{library}\NormalTok{(ggplot2)}

\CommentTok{\# custom functions}

\CommentTok{\# show first and last rows of a dataframe}
\NormalTok{head\_tail }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(x, }\AttributeTok{n =} \DecValTok{5}\NormalTok{)\{}
  \FunctionTok{rbind}\NormalTok{(}
    \FunctionTok{head}\NormalTok{(x, }\AttributeTok{n =}\NormalTok{ n),}
    \FunctionTok{tail}\NormalTok{(x, }\AttributeTok{n =}\NormalTok{ n)}
\NormalTok{  )}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

@@

We can generally define as \emph{ordinal regression}, a statistical model explicitly considering the ordinal (and not metric) nature of the response variable. The nomenclature of \emph{ordinal regression} models can be confusing but Tutz (2022) and Bürkner and Vuorre (2019) provide a clear and updated taxonomy of ordinal regression models.

We can identify three main models: \emph{cumulative models} (Agresti, 2010; CM, McCullagh, 1980), \emph{sequential models} (Tutz, 1990), and \emph{adjacent category models}. Among these, the CM is the most widely used, assuming the existence of a latent variable categorized using a set of thresholds. The \emph{sequential model} is appropriate when modeling sequential processes where responding to a certain category implies have already reached lower categories. The \emph{adjacent category model} compares a category with the successive one and can be seen the adjacent-category model can be seen as a series of binomial regressions taking into account the ordering (Tutz, 2022).

In the current paper we focus on the CM for several reasons. Firstly, the latent formulation of the model is particularly convenient for parameter interpretation and data simulation. The second reason is that several psychological construct can be formalized as a latent continuous variable observed as an ordinal item(s). Furthermore, the signal detection theory framework that is very common in experimental psychology can be implemented using a CM (e.g., DeCarlo, 2010). Figure \ref{fig:fig-explain-cumulative} depicts the overall structure of the CM.

@@

In this section we introduce some notation for the CM to be consistent with the existing literature and the \texttt{ordinal} R package (Christensen, 2022). We define \(Y_k\) as the observed ordinal variable with \(k\) levels and \(Y^\star\) is the underlying latent variable. The latent variable is segmented using \(k - 1\) thresholds \(\alpha_1, \dots, \alpha_{k - 1}\). Similarly to the generalized linear models framework (e.g., Fox, 2015), we define \(g(p) = \eta\) as the link function that maps probabilities into the linear predictor \(\eta\). To transform back \(\eta\) into probabilities we use the inverse of the link function \(p = g^{-1}(\eta)\). The specific link function define the type of model and require different R functions. For example, using a Normal distribution (\emph{probit} link function) requires using the cumulative distribution function \(g(p) = \Phi^{-1}(p) = \eta\) and the inverse of the link function is the inverse cumulative distribution function (or quantile function) defined \(p = g^{-1}(\eta) = \Phi(\eta)\).

When modelling an ordinal variable in a CM we include predictors on the cumulative probability \(P(Y \leq k), k = 1, \dots, k - 1\)\footnote{As done by Agresti (2010), when referring to \(P(Y \leq k)\) we are implicitly conditioning on a particular \(\mathbf{X}\) value \(P(Y \leq k | \mathbf{X}_i)\)}. Equation \eqref{eq:prob-cum-model1} shows the general cumulative model with predictors \(\mathbf{X}\) and coefficients \(\boldsymbol{\beta}\). The minus sign in \(\mathbf{X} \boldsymbol{\beta}\) is used to interpret the \(\beta_j\) as in the standard regression models where higher positive values corresponds to increased probability of responding higher \(k\) categories (Agresti, 2010).

@@

The \(\mathbf{X} \boldsymbol{\beta}\) is the linear predictor \(\eta\) that is the cumulative probability \(P(Y \leq k)\) transformed using the link function \(g(\cdot)\). The probability of a single outcome \(P(Y = k)\) can be calculated as the difference between cumulative probabilities (see Equation \eqref{eq:prob-cum-model2}).

@@

There are always two special cases when computing the probability of a single outcome \(Y\) that is when \(Y = 1\) and \(Y = k\). In the first case the cumulative probability is calculated from \(-\infty\) (assuming that \(\alpha_0 = -\infty\)). In the second case (\(Y = 1\)) the probability is calculated as \(P(Y = k) = 1 - g^{-1}(\alpha_{k - 1} - \eta)\) (a \(\alpha_k = +\infty\). The Figure \ref{fig:fig-explain-cumulative} shows how the single probabilities of the ordinal outcome are calculated from cumulative probabilities.

@@

\begin{equation}
\mathbf{Y^\star} = \mathbf{X}\boldsymbol{\beta} + \boldsymbol{\epsilon}
\label{eq:latent-model}
\end{equation}

The \(\epsilon\) term is the random part (errors) of the model. For a \emph{probit} model, errors are distributed as a standard Normal distribution and for a \emph{logit} model as a standard logistic distribution. Following the notation by Tutz (2022), the observed ordinal value \(Y_i = k\) comes from \(Y^\star_i\) belonging to the interval defined by the thresholds \(Y_i = k \iff \alpha_{k - 1} < Y^\star_i < \alpha_{k}\) where \(- \infty = \alpha_0 < \alpha_1 < \dots< \alpha_{k - 1} < \alpha_k = \infty\).

In the CM, thresholds \(\alpha_k\) are considered fixed and part of the measurement procedure (Liddell \& Kruschke, 2018; but see the \emph{location-shift} models by Tutz, 2022, where thresholds vary as a function of predictors) and can be considered as intercepts. The proposed CM can be parametrized in different ways. Liddell and Kruschke (2018) and Kruschke (2015) proposed a Bayesian version of the model and Gelman, Hill, and Vehtari (2020) proposed other thresholds parametrizations.

@@

The CM implemented in Equations \eqref{eq:prob-cum-model1} and \eqref{eq:latent-model} requires specifying the link function \(g(\cdot)\) or the errors distribution \(\epsilon_i \sim D(\mu, \sigma^2)\) (\(D\) being a certain probability distribution). Among several available functions the \emph{logit} and \emph{probit} models are the most common. The \emph{logit} model use a \emph{logit} link function and a standard logistic distribution as latent variable. On the other side, the \emph{probit} model assume a standard Normal distribution using the inverse of the cumulative distribution (\(\Phi^{-1}(\cdot)\)) function as link function.

@@

In terms of parameters, both distributions can be defined with in terms of location \(\mu\) and scale \(s\). The standard normal distribution has \(\mu = 0\) and \(s = 1\). Furthermore the variance corresponds to the scale \(s^2 = \sigma^2 = 1\). The variance of the logistic distribution is \(\sigma^2 = \frac{s^2\pi^2}{3}\). The standard logistic distribution has \(\mu = 0\) and \(s^2 = 1\) thus the standard deviation simplified to \(\frac{\pi}{\sqrt{3}} \approx 1.81\). In practical terms, fixing \(\mu\) and \(s\) lead to an higher standard deviation for the logistic distribution. In the supplementary materials we showed how to implement a location-scale models for including predictors on the scale parameter (Cox, 1995; Rigby \& Stasinopoulos, 2005; Tutz, 2022).

@@

For fitting the CM we used the \texttt{ordinal} package (Christensen, 2019). The \texttt{ordinal} package provide a complete and intuitive way to implement the ordinal models. The syntax is very similar to standard linear models in R with functions to calculate predictions, perform model comparison and extract relevant model information.\footnote{For a very complete overview of the ordinal package see Christensen (2019) and the Package documentation \url{https://cran.r-project.org/web/packages/ordinal/ordinal.pdf}}

@@

When fitting the model the crucial arguments are the \texttt{formula}, the \texttt{link} function and the \texttt{data}. More advanced arguments are \texttt{nominal}, \texttt{scale} and \texttt{threshold}.

\begin{itemize}
\tightlist
\item
  \texttt{formula}: the formula \texttt{y\ \textasciitilde{}\ x} with the dependent variable and predictors.
\item
  \texttt{link}: is the link function (in this tutorial only \emph{logit} and \emph{probit})
\item
  \texttt{data}: is the dataset with variables included in the \texttt{formula}
\item
  \texttt{nominal}: formula with predictors where the proportional odds assumption (See Section \ref{proportional-odds-assumption}) is relaxed (i.e., partial or non proportional odds)
\item
  \texttt{scale}: formula with predictors for the scale parameter for fitting a scale-location model
\item
  \texttt{threshold}: different structures for estimating the thresholds. The default is \texttt{threshold\ =\ "flexible"} where \(k - 1\) threshold (where \(k\) is the number of ordinal levels for \(Y\)) are estimated.
\end{itemize}

We can start by fitting a simple model, highlighting the crucial parameters where the detailed explanation will be expanded in the next sections. Table \ref{tab:tab-dataset-example-latex} contains simulated data from \(n = 5000\) large \(n\) for having stable estimations in the examples) participants rating the agreement about an item with \(k = 4\) ordered options. The participants are divided into two groups (\(X_a\) and \(X_b\)). We can fit a cumulative link model (\texttt{clm()} function) and check the model summary.

@@

\begin{table}
\centering
\caption{\label{tab:tab-dataset-example-latex}Summary of the simulated dataset. For each group (a and b) we reported mean, median and standard deviation of the ordinal outcome (metric descriptive statistics). For each ordinal outcome (\(Y\)) we reported the frequency and (within parentheses) the probability.}
\centering
\begin{tabular}[t]{cccccccc}
\toprule
\textbf{Group} & \textbf{Mean} & \textbf{Median} & \textbf{SD} & \textbf{Y1} & \textbf{Y2} & \textbf{Y3} & \textbf{Y4}\\
\midrule
a & 2.5 & 2 & 1.1 & \textbf{599} (p = \textit{11.98}) & \textbf{662} (p = \textit{13.24}) & \textbf{627} (p = \textit{12.54}) & \textbf{612} (p = \textit{12.24})\\
b & 3.17 & 3.5 & 0.99 & \textbf{221} (p = \textit{4.42}) & \textbf{381} (p = \textit{7.62}) & \textbf{648} (p = \textit{12.96}) & \textbf{1250} (p = \textit{25})\\
\bottomrule
\end{tabular}
\end{table}

@@

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{fit }\OtherTok{\textless{}{-}} \FunctionTok{clm}\NormalTok{(y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ x, }\AttributeTok{data =}\NormalTok{ dat, }\AttributeTok{link =} \StringTok{"logit"}\NormalTok{)}
\FunctionTok{summary}\NormalTok{(fit)}
\CommentTok{\#\textgreater{} formula: y \textasciitilde{} x}
\CommentTok{\#\textgreater{} data:    dat}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{}  link  threshold nobs logLik   AIC      niter max.grad cond.H }
\CommentTok{\#\textgreater{}  logit flexible  5000 {-}6458.45 12924.89 4(0)  1.59e{-}08 1.7e+01}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Coefficients:}
\CommentTok{\#\textgreater{}    Estimate Std. Error z value Pr(\textgreater{}|z|)    }
\CommentTok{\#\textgreater{} xb  1.14864    0.05333   21.54   \textless{}2e{-}16 ***}
\CommentTok{\#\textgreater{} {-}{-}{-}}
\CommentTok{\#\textgreater{} Signif. codes:  0 \textquotesingle{}***\textquotesingle{} 0.001 \textquotesingle{}**\textquotesingle{} 0.01 \textquotesingle{}*\textquotesingle{} 0.05 \textquotesingle{}.\textquotesingle{} 0.1 \textquotesingle{} \textquotesingle{} 1}
\CommentTok{\#\textgreater{} }
\CommentTok{\#\textgreater{} Threshold coefficients:}
\CommentTok{\#\textgreater{}     Estimate Std. Error z value}
\CommentTok{\#\textgreater{} 1|2 {-}1.16300    0.04345 {-}26.765}
\CommentTok{\#\textgreater{} 2|3  0.01284    0.03835   0.335}
\CommentTok{\#\textgreater{} 3|4  1.14143    0.04194  27.214}
\end{Highlighting}
\end{Shaded}

@@

Odds and odds ratios are clearly defined with \(k = 2\) outcomes. With an ordinal variable we have \(k - 1\) odds ratios determined by the cumulative probability in terms of \(P(Y \leq 1), \dots, P(Y \leq k - 1)\) comparing the two groups.

For example the probability of responding \(Y \leq 1\) in the group ``a'' is 0.24 corresponding to an odds of 0.32. When calculating the odds ratio comparing ``a'' vs ``b'' for \(Y \leq 1\) we obtain that the group ``a'' has 3.25 times the odds of responding \(Y \leq 1\) compared to the group ``b''.

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# the dummy\_ord() is a custom function creating k {-} 1 dummy variable for the cumulative probabilities}

\NormalTok{odds }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(p) p }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{{-}}\NormalTok{ p)}

@@

\CommentTok{\# calculating the cumulative probability for k {-} 1 variables. }
\CommentTok{\# Taking the average of a series of 0{-}1 is the same as computing the proportion of 1s.}
\NormalTok{(group\_a }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"a"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
\CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
\CommentTok{\#\textgreater{}  0.2396  0.5044  0.7552}
\NormalTok{(group\_b }\OtherTok{\textless{}{-}} \FunctionTok{apply}\NormalTok{(cum\_p[cum\_p}\SpecialCharTok{$}\NormalTok{x }\SpecialCharTok{==} \StringTok{"b"}\NormalTok{, }\SpecialCharTok{{-}}\DecValTok{1}\NormalTok{], }\DecValTok{2}\NormalTok{, mean))}
\CommentTok{\#\textgreater{} y1vs234 y12vs34 y123vs4 }
\CommentTok{\#\textgreater{}  0.0884  0.2408  0.5000}
